---
title: "Assignment 3: Physical Properties of Rivers"
author: Yingfan Zeng
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on the physical properties of rivers. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/ydeD5axzCnaNzgss9

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-02-08

## Setup

1. Verify your working directory is set to the R project file. Load the tidyverse, dataRetrieval, lubridate, and lfstat packages. Set your ggplot theme (can be theme_classic or something else).
2. Import a data frame called "MysterySiteDischarge" from USGS gage site 03431700. Import discharge data starting on 1964-10-01 and ending on 2021-09-30. Rename columns 4 and 5 as "Discharge" and "Approval.Code". DO NOT LOOK UP WHERE THIS SITE IS LOCATED. 
3. Build a ggplot of discharge over the entire period of record. 

```{r}
getwd()

#Load packages
library(tidyverse)
library(lubridate)
library(dataRetrieval)
library(lfstat)

#Set ggplot theme
MyTheme <- theme_bw(base_size = 14) +
  theme(axis.text = element_text(color = "black"),
        legend.position = "right",
        plot.title = element_text(hjust = 0.5))

#Import dataset
MysterySiteDischarge <- readNWISdv(siteNumbers = "03431700",
                     parameterCd = c("00060", "00065"),
                     startDate = "1964-10-01",
                     endDate = "2021-09-30")

#Rename columns
names(MysterySiteDischarge)[4:5] <- c("Discharge", "Approval.Code")

#Build a ggplot
ggplot(MysterySiteDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  MyTheme

```

## Analyze seasonal patterns in discharge

4. Add a "WaterYear" and "DayOfYear" column to the data frame. Hint: Use a pipe, and you will need both the lubridate and lfstat packages. Set WaterYear to numeric.
5. Create a new data frame called "MysterySiteDischarge.Pattern" that has columns for Day.of.Year, median discharge for a given day of year, 75th percentile discharge for a given day of year, and 25th percentile discharge for a given day of year. Hint: the summarise function includes `quantile`, wherein you must specify `probs` as a value between 0 and 1.
6. Create a plot of median, 75th quantile, and 25th quantile discharges against day of year. Median should be black, other lines should be gray. 
```{r}
#Add columns of water year and DOY
MysterySiteDischarge <- MysterySiteDischarge %>%
  mutate(WaterYear = as.numeric(as.character(water_year(Date, origin = "usgs"))),
         DayOfYear = yday(Date))

#Calculate median, Q75 and Q25
MysterySiteDischarge.Pattern <- MysterySiteDischarge %>%
  group_by(DayOfYear) %>%
  summarise(median = median(Discharge),
            Q75 = quantile(Discharge, probs = 0.75),
            Q25 = quantile(Discharge, probs = 0.25))

#Plot them
ggplot(MysterySiteDischarge.Pattern, aes(x = DayOfYear)) +
  geom_line(aes(y = median), color = "black") +
  geom_line(aes(y = Q75), color = "gray") +
  geom_line(aes(y = Q25), color = "gray") +
  MyTheme +
  labs(y = "Discharge (cfs)", x = "DOY")

```

7. What seasonal patterns do you see? What does this tell you about precipitation patterns and climate in the watershed?

> This site has high discharge in winter and spring, while low discharge in the summer. It indicates that there is high precipitation in winter and small precipitation in summer in this watershed. 

## Create and analyze recurrence intervals

8. Create two separate data frames for MysterySite.Annual.30yr (first 30 years of record) and MysterySite.Annual.Full (all years of record). Use a pipe to create your new data frame(s) that includes the water year, the peak discharge observed in that year, a ranking of peak discharges, the recurrence interval, and the exceedende probability.

9. Create a plot that displays the discharge vs. recurrence interval relationship for the two separate data frames (one set of points includes the values computed from the first 30 years of the record and the other set of points includes the values computed for all years of the record. 

10. Create a model to predict the discharge for a 100-year flood for both sets of recurrence intervals. 

```{r}
#Calculate the probability of recurrence for the first 30 years
MysterySite.Annual.30yr <- MysterySiteDischarge %>%
  subset(WaterYear <= 1995) %>%
  group_by(WaterYear) %>%
  summarise(PeakDischarge = max(Discharge)) %>%
  mutate(Rank = rank(-PeakDischarge), 
         RecurrenceInterval = (length(WaterYear) + 1)/Rank, 
         Probability = 1/RecurrenceInterval)

#Calculate the probability of recurrence for all years
MysterySite.Annual.Full <- MysterySiteDischarge %>%
  group_by(WaterYear) %>%
  summarise(PeakDischarge = max(Discharge)) %>%
  mutate(Rank = rank(-PeakDischarge), 
         RecurrenceInterval = (length(WaterYear) + 1)/Rank, 
         Probability = 1/RecurrenceInterval)

#Plot them
ggplot() +
  geom_point(MysterySite.Annual.30yr, mapping = aes(x = PeakDischarge, y = RecurrenceInterval, col = "first 30 yr")) +
  geom_point(MysterySite.Annual.Full, mapping = aes(x = PeakDischarge, y = RecurrenceInterval, col = "full record")) +
  scale_x_log10() +
  MyTheme

#Create a model to predict the discharge for a 100-year flood (30 yr)
floodModel.30yr <- lm(data = MysterySite.Annual.30yr, PeakDischarge ~ log10(RecurrenceInterval))
summary(floodModel.30yr)
floodModel.30yr$coefficients

floodModel.30yr$coefficients[1] + floodModel.30yr$coefficients[2]*log10(100)
#5515.593

#Create a model to predict the discharge for a 100-year flood (full record)
floodModel.full <- lm(data = MysterySite.Annual.Full, PeakDischarge ~ log10(RecurrenceInterval))
summary(floodModel.full)
floodModel.full$coefficients

floodModel.full$coefficients[1] + floodModel.full$coefficients[2]*log10(100)
#4835.488
```

11. How did the recurrence interval plots and predictions of a 100-year flood differ among the two data frames? What does this tell you about the stationarity of discharge in this river?

> A 100-year flood predicted by the 30-year record is 4835 cfs while it is 5516 cfs predicted by the full record. From the plot, high discharge (>1000 cfs) has a higher recurrence interval than predicted by the first 30-year of record. It means according to the full record, there is a smaller chance for high discharge to happen. These mean the discharge at this site decreased after 1995.


