---
title: "Assignment 4: Physical Properties of Rivers"
author: Yingfan Zeng
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on the physical properties of rivers. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/futQwtCsyYsZG9nCA

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-02-15

## Setup and Data Processing

1. Verify your working directory is set to the R project file. Load the tidyverse, cowplot, dataRetrieval, lubridate, lfstat, and EcoHydRology packages. Set your ggplot theme (can be theme_classic or something else).
2. Acquire daily mean discharge data for the Bitterroot River in Montana (USGS gage 12344000) and the Nehalem River in Oregon (USGS gage 14299800). Collect the 10 most recent complete water years.
3. Add columns in the dataset for water year, baseflow, and stormflow. Feel free to use either baseflow separation function we used in class. 
4. Calculate annual total discharge, annual baseflow, and annual proportion baseflow for the two sites.

```{r}
getwd()

#Load packages
library(EcoHydRology)
library(tidyverse)
library(cowplot)
library(dataRetrieval)
library(lubridate)
library(lfstat)


#Set ggpplot theme
MyTheme <- theme_bw(base_size = 14) +
  theme(axis.text = element_text(color = "black"),
        legend.position = "right",
        plot.title = element_text(hjust = 0.5))

```

```{r}
#Import daily mean discharge data for the Bitterroot River in Montana
BittDischarge <- readNWISdv(siteNumbers = "12344000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2011-10-01", 
                     endDate = "2021-09-30")

names(BittDischarge)[4:5] <- c("Discharge", "Approval.Code")

#Add water year, baseflow, and stormflow
BittDischarge <- BittDischarge %>%
  mutate(WaterYear = as.numeric(as.character(water_year(Date))),
         Baseflow = baseflow(Discharge), 
         Stormflow = Discharge - Baseflow)

#Calculate annual total discharge, annual baseflow, and annual proportion baseflow
BittSummary <- BittDischarge %>%
  group_by(WaterYear) %>%
  summarise(Discharge.acft.yr = sum(Discharge)*723.968, # convert to ac*ft/yr
            Baseflow.acft.yr = sum(Baseflow)*723.968,
            Stormflow.acft.yr = sum(Stormflow)*723.968,
            Prop.bf = Baseflow.acft.yr/Discharge.acft.yr)
            

```

```{r}
#Import daily mean discharge data for the Nehalem River in Oregon 
NehalemDischarge <- readNWISdv(siteNumbers = "14299800",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2011-10-01", 
                     endDate = "2021-09-30")

names(NehalemDischarge)[4:5] <- c("Discharge", "Approval.Code")

#Add water year, baseflow, and stormflow
NehalemDischarge <- NehalemDischarge %>%
  mutate(WaterYear = as.numeric(as.character(water_year(Date))),
         Baseflow = baseflow(Discharge), 
         Stormflow = Discharge - Baseflow)

#Calculate annual total discharge, annual baseflow, and annual proportion baseflow
NehalemSummary <- NehalemDischarge %>%
  group_by(WaterYear) %>%
  summarise(Discharge.acft.yr = sum(Discharge)*723.968, # convert to ac*ft/yr
            Baseflow.acft.yr = sum(Baseflow)*723.968,
            Stormflow.acft.yr = sum(Stormflow)*723.968,
            Prop.bf = Baseflow.acft.yr/Discharge.acft.yr)
            
```


## Analyze seasonal patterns in discharge

5. For both sites, create a graph displaying discharge and baseflow by date. Adjust axis labels accordingly.
6. For both sites, create a graph displaying annual total discharge and annual baseflow across years, and a second graph displaying the proportion baseflow across years (adjust axis labels accordingly). Plot these graphs on top of one another using `plot_grid`. Remember to align the axes!

```{r}
#Bitterroot River
ggplot(BittDischarge, aes(x = Date)) +
  geom_line(aes(y = Discharge, col = "Discharge")) +
  geom_line(aes(y = Baseflow, col = "Baseflow")) +
  MyTheme +
  labs(x = "Year", y = "Discharge (cfs)")

#6.1
Plot61 <- ggplot(BittSummary, aes(x = WaterYear)) +
  geom_line(aes(y = Discharge.acft.yr, col = "Discharge")) +
  geom_line(aes(y = Baseflow.acft.yr, col = "Baseflow")) +
  MyTheme +
  labs(x = "Year", y = "Discharge (cfs)") +
  theme(legend.position = "top")

#6.2
Plot62 <- ggplot(BittSummary, aes(x = WaterYear, y = Prop.bf)) +
  geom_col() +
  MyTheme +
  labs(x = "Year", y = "Proportion of baseflow")

#Combine
plot_grid(Plot61, Plot62, nrow = 2, align = "hv")

```

```{r}
#Bitterroot River
ggplot(NehalemDischarge, aes(x = Date)) +
  geom_line(aes(y = Discharge, col = "Discharge")) +
  geom_line(aes(y = Baseflow, col = "Baseflow")) +
  MyTheme +
  labs(x = "Year", y = "Discharge (cfs)")

#6.1
NYear <- ggplot(NehalemSummary, aes(x = WaterYear)) +
  geom_line(aes(y = Discharge.acft.yr, col = "Discharge")) +
  geom_line(aes(y = Baseflow.acft.yr, col = "Baseflow")) +
  MyTheme +
  labs(x = "Year", y = "Discharge (cfs)") +
  theme(legend.position = "top")

#6.2
NProp <- ggplot(NehalemSummary, aes(x = WaterYear, y = Prop.bf)) +
  geom_col() +
  MyTheme +
  labs(x = "Year", y = "Proportion of baseflow")

#Combine
plot_grid(NYear, NProp, nrow = 2, align = "hv")


```

7. How do these rivers differ in their discharge and baseflow, both within and across years? How would you attribute these patterns to the climatic conditions in these locations?

> The baseflow of the Bitterroot River follows closely with the discharge, especially in the winter. In summer, there were a lot of precipitation that excessed the baseflow and was considered as storm flow. This pattern is true for most of the years, while the winter between 2018 and 2019 had some discfharge peaks. The winter in Montana probably has less precipitation than in summer. 
For the Nehalem River, the baseflow is almost the same as the discharge in winters. However, it has active storm events in summer that was not conted as baseflow. This aligns with the climate in Oregon with massive summer rains. 
